Evaluation of Pre-Trained Models for Many-to-English Translation

### 2024-08-05 paper/paper.tex
updated and added result

% Packages
\usepackage{graphicx}
\usepackage[margin=1in]{geometry}
\usepackage[backend=bibtex]{biblatex}
\usepackage{comment}
\usepackage{hyperref}
\usepackage{amsmath}

@@

Through simple experimentations, this paper aims to evaluate the performance of existing pre-trained models (PTMs) on many-to-English translation across 14 source languages. Although fine-tuning multilingual PTMs has been proven to increase model performance \cite{cooper-stickland-2021-recipes}, no pre-training or fine-tuning is performed in this study for simplicity reasons. A dataset is curated from the Tatoeba repository \cite{tatoeba}, containing 1,241 parallel sentence pairs across source and target languages. The models includes a one-to-one PTMs: OPUS-MT \cite{tiedemann-2020-opus-mt}, and multilingual PTMs: such as mBART-50 \cite{liu-2020-mbart}, NLLB-200 \cite{nllb200-2020}, and M2M-100 \cite{fan-2020-m2m100}. The performance of these models is evaluated using the BLEU score \cite{papieni-2002-bleu}.

@@

Since Neural Machine Translation (NMT) systems require vast amounts of training data, the availability of parallel corpora is crucial for building effective models \cite{koehn-2017-challenges}. The lack of extensive parallel corpora, especially for low-resource languages, leads to suboptimal performance in NMT techniques compared to their high-resource counterparts \cite{ranathunga-2023-nmt-low-res}. OPUS \cite{opus} is a comprehensive collection of open-source parallel corpora used extensively in the field of machine translation (MT). It includes corpora for 744 languages and contains over 1,210 different datasets, amassing a total of 45,945,946,108 sentence pairs. Tatoeba \cite{tatoeba} is another prominent resource in the field of MT and NLP, known for its extensive collection of translated sentences. As of July 2024, it contains 12,186,207 sentences over 423 supported languages, growing daily through volunteer contributions.

The Tatoeba Challenge \cite{tiedemann-2020-tatoeba-challenge}, IWSLT \cite{agarwal-iwstl-2023}, TED \cite{ye-2018-word-embeddings-ted}, Flores \cite{goyal-2022-flores}, and WMT \cite{barrault-2020-wmt} are among the biggest and most popular parallel corpora datasets, allowing for many-to-many translations.

Despite being used diligently by Meta AI papers for benchmarking \cite{fan-2020-m2m100,nllb200-2020}, these datasets are not seen used

@@

One-to-One Translation refers to a translation approach where a model is specifically trained to translate between one source language and one target language. This setup is characterised by having a dedicated model for each unique language pair. An example of this setup is OPUS-MT by Helsinki-NLP \cite{tiedemann-2020-opus-mt}, which provides over 1,000 pre-trained models for translation between numerous language pairs.

@@

mRASP2 \cite{pan-2021-mrasp2}

@@

@@

While other metrics exist for machine translation such as METEOR \cite{lavie-2007-meteor} and COMET \cite{rei-2020-comet}, it is far more practical to implement BLEU due to its popular usage in other works.

% Introduce the concept of evaluation metrics for MT, explaining why BLEU is chosen for this study and if there are other metrics that could complement it (like METEOR or ROUGE).

% \begin{comment}
% PolyLM \cite{wei-2023-polylm} (32 languages), and
% mT5 \cite{xue-2021-mt5} (101 languages).

% \subsection{Sequence-to-Sequence Models}

% GPT \cite{openai-2024-gpt4}

% mBERT \cite{devlin-2019-bert} is the multilingual version of BERT, introduced in the same paper, trained on 104 different languages. However, the model been shown to suffer on low-resource languages \cite{wu-2020-mbert-are-all}.

% T5 \cite{raffel-2023-t5} is a Text-to-Text Transfer Transformer model trained on English.

% MistralAI \cite{jiang-2023-mistral-7b}

% \end{comment}

@@

\begin{table}[htbp]
    \centering
    \begin{tabular}{|l|l|}
        \hline
        \textbf{No.} & \textbf{Language} \\
        \hline
        1            & Dutch             \\
        2            & Finnish           \\
        3            & French            \\
        4            & German            \\
        5            & Hebrew            \\
        6            & Hungarian         \\
        7            & Italian           \\
        8            & Japanese          \\
        9            & Mandarin Chinese  \\
        10           & Polish            \\
        11           & Russian           \\
        12           & Spanish           \\
        13           & Turkish           \\
        14           & Ukrainian         \\
        \hline
    \end{tabular}
    \caption{List of chosen languages for evaluation}
    \label{table:eval_languages}
\end{table}

Table \ref{table:eval_languages} show the 14 languages selected for this project. Languages are chosen based on its resources' availability in Tatoeba, as well as considering supported languages in most PTMs models. To build the dataset, sentences in English are first downloaded, containing 1,898,494 sentences (it is unclear why it is less than the number stated in the Tatoeba website). Then for each language, sentence pairs between English and source languages are downloaded individually and compiled. The result is a single Dataframe containing 1241 sentences in all 14 languages, this will be treated as a test set to evaluate the models performance on each language.

Sentences typically consist of everyday phrases such as 'I have to go to sleep', 'That is intriguing', and 'Where do you live?' They may also include single-word exclamations like 'Speak!' or 'Look!' Additionally, multiple sentences such as 'You may write in any language you want. On Tatoeba, all languages are considered equal', and 'Guns don't kill people. People kill people' can be found inside the corpus. A few of them also include human names, 'Compare your answer with Tom's', 'Muiriel is 20 now'. All of the sentences are straightforward and literal, without the use of linguistic devices such as metaphors or sarcasm. Therefore, machine translation process should be straightforward on this level.

@@

>>>>

% Packages
\usepackage{graphicx}
\usepackage[margin=1in]{geometry}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[backend=bibtex]{biblatex}
\usepackage{comment}
\usepackage{hyperref}
\usepackage{amsmath}

\usepackage{CJKutf8}

@@

Through simple experimentations, this paper aims to evaluate the performance of existing pre-trained models (PTMs) on many-to-English translation across 14 source languages. Although fine-tuning multilingual PTMs has been proven to increase model performance \cite{cooper-stickland-2021-recipes}, no pre-training or fine-tuning is performed in this study for simplicity reasons. A dataset is curated from the Tatoeba repository \cite{tatoeba}, containing 1,323 parallel sentence pairs across source and target languages. The models includes a one-to-one PTMs: OPUS-MT \cite{tiedemann-2023-democratizing,tiedemann-2020-opus-mt}, and multilingual PTMs: such as mBART-50 \cite{liu-2020-mbart}, NLLB-200 \cite{nllb200-2020}, and M2M-100 \cite{fan-2020-m2m100}. The performance of these models is evaluated using the BLEU score \cite{papieni-2002-bleu}.

@@

@@

One-to-One Translation refers to a translation approach where a model is specifically trained to translate between one source language and one target language. This setup is characterised by having a dedicated model for each unique language pair. An example of this setup is OPUS-MT by Helsinki-NLP \cite{tiedemann-2023-democratizing,tiedemann-2020-opus-mt}, which provides over 1,000 pre-trained models for translation between numerous language pairs.

@@

Corpora are large and structured sets of texts used for linguistic research and analysis. Thus, parallel corpora are defined as sets of texts in a given source language along with their translations in another target language \cite{lefer-2020-parallel-corpora}. It can be bilingual or multilingual, and are crucial for training and evaluating machine translation. Languages without or with limited parallel corpora are referred as \textbf{low-resources language}.

Since the early 2000s, English has often been used as an intermediary language, meaning texts originally written in languages are first translated into English, and then from English into other languages, resulting in most target texts being translations of translations \cite{lefer-2020-parallel-corpora}. This is called indirect translation, and posed for many problems in translations as terms can be ambiguous with varying or similar meanings \cite{rosa-2017-indirect-translation-problems}.

Neural Machine Translation (NMT) systems require vast amounts of training data, and thus the availability of parallel corpora is crucial for building effective models \cite{koehn-2017-challenges}. The lack of extensive parallel corpora, especially for low-resource languages, leads to suboptimal performance in NMT techniques compared to their high-resource counterparts \cite{ranathunga-2023-nmt-low-res}. OPUS \cite{opus} is a comprehensive collection of open-source parallel corpora used extensively in the field of machine translation (MT). It includes corpora for 744 languages and contains over 1,210 different datasets, amassing a total of 45,945,946,108 sentence pairs. Tatoeba \cite{tatoeba} is another prominent resource in the field of MT and NLP, known for its extensive collection of translated sentences. As of July 2024, it contains 12,186,207 sentences over 423 supported languages, growing daily through volunteer contributions.

Several datasets are often used for evaluation in machine translation. The Tatoeba Challenge \cite{tiedemann-2020-tatoeba-challenge} covers 487 languages in 4,024 language pairs, including 657 test sets sourced from Tatoeba website, covering 138 languages. The TED Talks dataset \cite{ye-2018-word-embeddings-ted} contains transcripts from TED talks for more than 50 languages. IWSLT \cite{agarwal-iwstl-2023} also contains TED talks data, but paired with English translations. Finally, WMT \cite{barrault-2020-wmt} is an annual event that organises tasks for machine translation and provides a collection of datasets for benchmarking and evaluating translation systems.

While these datasets can be used to evaluate translation performance between any language pairs, they do not contain parallel corpora between different languages. Corpus between English and French for example, contains different text to the corpus between English and Spanish. Therefore, they do not support easy evaluation for many-to-English translations.

Furthermore, papers often use different sets and different versions of datasets to evaluate their model performance, incurring unstraightforward comparison.

mBart \cite{liu-2020-mbart} uses WMT19 for English-German translation and TED15 for Chinese-English translation. M2M-100 \cite{fan-2020-m2m100} uses 7 different datasets shared across languages. NLLB-200  \cite{nllb200-2020} uses FLORES dataset, designed for low-resources languages. OPUS-MT \cite{tiedemann-2023-democratizing,tiedemann-2020-opus-mt} models use datasets from Tatoeba challenges.

@@

Other metrics for machine translation exist, but not used nearly as widely as BLEU. Crosslingual Optimized Metric for Evaluation of Translation (COMET) \cite{rei-2020-comet} is a recent metric that leverages neural-network-based models for deeper understanding and evaluation. BERTScore.

In this project, a variant of BLEU called SacreBLEU and METEOR will be used due to their simplicity.

\subsubsection{BLEU}

@@

The machine translation community's rely heavily BLEU score, however, it has several drawbacks. The metric has been reported to not correlate strongly with human judgement, showing variations in translation that could mean that a higher BLEU score does not necessarily indicate a true enhancement in translation quality \cite{callison-burch-2006-reevaluating-bleu}.

Furthermore, It is challenging to directly compare BLEU scores between paper \cite{post-2018-clarity-bleu}. Thus, the author proposed a variation called SacreBLEU \cite{post-2018-clarity-bleu}, which claimed to facilitate easy computation of BLEU scores that are shareable, comparable, and reproducible. This is done by including preprocessing steps and normalisation routines.

\subsubsection{METEOR}

Metric for Evaluation of Translation with Explicit ORdering (METEOR) \cite{lavie-2007-meteor} assesses a translation by calculating a score that reflects explicit word-to-word matches between the reference and a candidate translation \cite{agarwal-2008-meteor-mbleu-mter}. It is designed to address some limitations of the BLEU score, allowing matches between simple morphological variants and synonyms.

\begin{equation}
    \text{METEOR} = (1 - \gamma \cdot \text{frag}) \cdot \frac{P \cdot R}{\alpha \cdot P + (1 - \alpha) \cdot R}
\end{equation}

Where:
\begin{itemize}
    \item \(P\) is the precision,
    \item \(R\) is the recall,
    \item \(\text{frag}\) is the fragmentation penalty,
    \item \(\gamma\) is a parameter that controls the weight of the fragmentation penalty, commonly 0.5,
    \item \(\alpha\) is a parameter that controls the balance between precision and recall, commonly 0.9,
\end{itemize}

@@

\begin{table}[htbp]
    \centering
    \begin{tabular}{|l|l|r|}
        \hline
        \textbf{No.} & \textbf{Language} & \textbf{Sentence Pairs Count} \\
        \hline
        1            & Chinese           & 68,814                        \\
        2            & Dutch             & 155,856                       \\
        3            & Finnish           & 102,202                       \\
        4            & French            & 405,088                       \\
        5            & German            & 501,145                       \\
        6            & Hebrew            & 172,082                       \\
        7            & Hungarian         & 171,698                       \\
        8            & Italian           & 624,160                       \\
        9            & Japanese          & 270,116                       \\
        10           & Polish            & 77,345                        \\
        11           & Russian           & 722,837                       \\
        12           & Spanish           & 265,253                       \\
        13           & Turkish           & 710,279                       \\
        14           & Ukrainian         & 214,244                       \\
        \hline
    \end{tabular}
    \caption{List of chosen languages for evaluation}
    \label{table:eval_languages}
\end{table}

Table \ref{table:eval_languages} show the 14 languages selected for this project. Languages are chosen based on its resources' availability in Tatoeba, as well as considering supported languages in most PTMs models. To build the dataset, sentences in English are first downloaded, containing 1,898,494 sentences (it is unclear why it is less than the number stated in the Tatoeba website). Then for each language, sentence pairs between English and source languages are downloaded individually and compiled. The result is a single Dataframe containing 1,323 parallel sentences in all 14 languages, this will be treated as a test set to evaluate the models performance on each language.

Sentences typically consist of everyday phrases such as 'I have to go to sleep', 'That is intriguing', and 'Where do you live?'. They may also include single-word exclamations like 'Speak!' or 'Look!'. Additionally, multiple sentences such as 'You may write in any language you want. On Tatoeba, all languages are considered equal', and 'Guns don't kill people. People kill people' can be found inside the corpus. A few of them also include human names, 'Compare your answer with Tom's', 'Muiriel is 20 now'. All of the sentences are straightforward and literal, without the use of linguistic devices such as metaphors or sarcasm. Therefore, machine translation process should be straightforward on this level.

\begin{CJK}{UTF8}{gbsn}
    \begin{table}[htbp]
        \centering
        \begin{tabular}{|l|l|l|}
            \hline
            \textbf{Language} & \textbf{Sentence 1}                   & \textbf{Sentence 2}                   \\
            \hline
            English           & I have to go to sleep.                & So what?                              \\
            Chinese           & 我该去睡觉了。                               & 那又怎樣?                                 \\
            Dutch             & Ik moet gaan slapen.                  & Dus?                                  \\
            Finnish           & Minun täytyy mennä nukkumaan.         & Mitä sitten?                          \\
            French            & Je dois aller dormir.                 & Et alors ?                            \\
            German            & Ich muss jetzt schlafen.              & Na und?                               \\
            Hebrew            & <hidden-due-to-latex-incompatibility> & <hidden-due-to-latex-incompatibility> \\
            Hungarian         & Aludni kell mennem.                   & És akkor mi van?                      \\
            Italian           & Devo andare a dormire.                & E allora?                             \\
            Japanese          & 私は眠らなければなりませ                          & だから何？                                 \\
            Polish            & Muszę iść spać.                       & No i co?                              \\
            Russian           & Мне пора идти                         & Так что?                              \\
            Spanish           & Tengo que irme a dormir.              & ¿Entonces qué?                        \\
            Turkish           & Yatmaya gitmek zorundayım.            & Öyleyse ne yapmalı?                   \\
            Ukrainian         & Маю піти спати.                       & Ну то що?                             \\
            \hline
        \end{tabular}
        \caption{A snippet of the dataset}
        \label{table:parallel_sentence}
    \end{table}
\end{CJK}

Figure \ref{table:parallel_sentence} shows a few examples of parallel sentences in the final dataset.

@@

\begin{table}[htbp]
    \centering
    \begin{minipage}{0.49\linewidth}
        \footnotesize
        \begin{tabular}{|l|r|r|r|}
            \hline
            \textbf{Language} & \textbf{BLEU} & \textbf{SacreBLEU} & \textbf{METEOR} \\
            \hline
            Chinese           & 0.0975        & 43.3490            & 0.6848          \\
            Dutch             & 0.0976        & 52.6920            & 0.7448          \\
            Finnish           & 0.0975        & 49.9149            & 0.7145          \\
            French            & 0.0969        & 51.4090            & 0.7288          \\
            German            & 0.0969        & 52.7877            & 0.7423          \\
            Hebrew            & 0.0981        & 50.3287            & 0.7198          \\
            Italian           & 0.0975        & 53.1644            & 0.7341          \\
            Japanese          & 0.0959        & 41.6610            & 0.6511          \\
            Polish            & 0.0984        & 52.5151            & 0.7301          \\
            Russian           & 0.0976        & 48.2725            & 0.7027          \\
            Spanish           & 0.0977        & 52.9323            & 0.7371          \\
            Turkish           & 0.0978        & 51.3009            & 0.7262          \\
            Ukrainian         & 0.0978        & 46.5555            & 0.6890          \\
            \hline
        \end{tabular}
        \caption{M2M-100}
    \end{minipage}
    \begin{minipage}{0.49\linewidth}
        \footnotesize
        \begin{tabular}{|l|r|r|r|}
            \hline
            \textbf{Language} & \textbf{BLEU} & \textbf{SacreBLEU} & \textbf{METEOR} \\
            \hline
            Chinese           & 0.0962        & 54.7322            & 0.7600          \\
            Dutch             & 0.0970        & 63.6482            & 0.8007          \\
            Finnish           & 0.0963        & 47.2194            & 0.7067          \\
            French            & 0.0968        & 57.2482            & 0.7598          \\
            German            & 0.0967        & 63.1666            & 0.8022          \\
            Hebrew            & 0.0972        & 58.0846            & 0.7590          \\
            Italian           & 0.0972        & 65.9415            & 0.8068          \\
            Japanese          & 0.0943        & 43.9547            & 0.7151          \\
            Polish            & 0.0978        & 62.7550            & 0.7923          \\
            Russian           & 0.0974        & 58.8820            & 0.7686          \\
            Spanish           & 0.0883        & 35.1593            & 0.7306          \\
            Turkish           & 0.0975        & 50.9377            & 0.6982          \\
            Ukrainian         & 0.0964        & 55.8637            & 0.7496          \\
            \hline
        \end{tabular}
        \caption{mBART50}
    \end{minipage}
\end{table}

\begin{table}[htbp]
    \centering
    \begin{minipage}{0.49\linewidth}
        \footnotesize
        \begin{tabular}{|l|r|r|r|}
            \hline
            \textbf{Language} & \textbf{BLEU} & \textbf{SacreBLEU} & \textbf{METEOR} \\
            \hline
            Chinese           & 0.0949        & 50.4454            & 0.7180          \\
            Dutch             & 0.0915        & 20.8988            & 0.2899          \\
            Finnish           & 0.0921        & 32.3926            & 0.4006          \\
            French            & 0.0950        & 32.1389            & 0.4338          \\
            German            & 0.0903        & 7.1436             & 0.1528          \\
            Hebrew            & 0.0897        & 2.3455             & 0.1412          \\
            Italian           & 0.0926        & 24.2702            & 0.3059          \\
            Japanese          & 0.0842        & 1.9569             & 0.0919          \\
            Polish            & 0.0937        & 38.4395            & 0.4739          \\
            Russian           & 0.0918        & 20.6734            & 0.2977          \\
            Spanish           & 0.0929        & 26.3313            & 0.3578          \\
            Turkish           & 0.0945        & 45.2721            & 0.6187          \\
            Ukrainian         & 0.0914        & 9.0340             & 0.1613          \\
            \hline
        \end{tabular}
        \caption{NLLB-200}
    \end{minipage}
    \begin{minipage}{0.49\linewidth}
        \footnotesize
        \begin{tabular}{|l|r|r|r|}
            \hline
            \textbf{Language} & \textbf{BLEU} & \textbf{SacreBLEU} & \textbf{METEOR} \\
            \hline
            Chinese           & 0.0971        & 59.5098            & 0.7953          \\
            Dutch             & 0.0974        & 69.8803            & 0.8471          \\
            Finnish           & 0.0965        & 66.6267            & 0.8296          \\
            French            & 0.0967        & 69.8185            & 0.8357          \\
            German            & 0.0969        & 69.7422            & 0.8419          \\
            Hebrew            & 0.0977        & 66.5149            & 0.8229          \\
            Italian           & 0.0979        & 74.1298            & 0.8584          \\
            Japanese          & 0.0960        & 63.1435            & 0.7893          \\
            Polish            & 0.0952        & 61.9026            & 0.8425          \\
            Russian           & 0.0974        & 66.7046            & 0.8179          \\
            Spanish           & 0.0979        & 71.4174            & 0.8463          \\
            Turkish           & 0.0980        & 72.6551            & 0.8460          \\
            Ukrainian         & 0.0978        & 75.4447            & 0.8667          \\
            \hline
        \end{tabular}
        \caption{OPUS-MT}
    \end{minipage}
\end{table}

\begin{figure}[htbp]
    \centering
    \includegraphics[width=1\linewidth]{figures/metrics_bar.png}
    \caption{BLEU, SacreBLEU, and METEOR scores of each model on each language.}
    \label{fig:result_visual}
\end{figure}

Figure \ref{fig:result_visual} shows the visualisation of all scores.
